---
title: "Code availability for the paper Mutant Huntingtin impairs neurodevelopment in human brain organoids through CHCHD2-mediated neurometabolic failure"
output: html_notebook
---

```{r}
#R version 4.2.2
library(Seurat) #version 5.0.2 
library(dplyr) #version 1.1.4
library(ggplot2) #version 3.5.0
```
# Data pre-processing
Read digital gene expression matrices (generated with Cell Ranger v7.1.0 and GRCh38-2020-A) and create Seurat objects for each sample keeping barcodes with at least 500 detected genes and genes detected in at least 5 cells
```{r}
WT_1= Read10X('../GEO_submission/cellranger_DGE/WT_1/') %>% CreateSeuratObject(min.cells = 5, min.features = 500) %>% subset(nCount_RNA > 1500)
WT_1$sample_id= 'WT_1'

WT_2= Read10X('../GEO_submission/cellranger_DGE/WT_2/') %>% CreateSeuratObject(min.cells = 5, min.features = 500) %>% subset(nCount_RNA > 1500)
WT_2$sample_id= 'WT_2'

WT_3= Read10X('../GEO_submission/cellranger_DGE/WT_3/') %>% CreateSeuratObject(min.cells = 5, min.features = 500) %>% subset(nCount_RNA > 1500)
WT_3$sample_id= 'WT_3'

HD_1= Read10X('../GEO_submission/cellranger_DGE/HD_1/') %>% CreateSeuratObject(min.cells = 5, min.features = 500) %>% subset(nCount_RNA > 1500)
HD_1$sample_id= 'HD_1'

HD_2= Read10X('../GEO_submission/cellranger_DGE/HD_2/') %>% CreateSeuratObject(min.cells = 5, min.features = 500) %>% subset(nCount_RNA > 1500)
HD_2$sample_id= 'HD_2'

HD_3= Read10X('../GEO_submission/cellranger_DGE/HD_3/') %>% CreateSeuratObject(min.cells = 5, min.features = 500) %>% subset(nCount_RNA > 1500)
HD_3$sample_id= 'HD_3'
```

Merge Seurat objects and add 'condition' metadata
```{r}
full_object = merge(WT_1, list(WT_2, WT_3, HD_1, HD_2, HD_3))

full_object$condition= NA
full_object$condition[full_object$sample_id %in% c('WT_1', 'WT_2', 'WT_3')] = 'WT/WT'
full_object$condition[full_object$sample_id %in% c('HD_1', 'HD_2', 'HD_3')] = '70Q/70Q'
remove(WT_1, WT_2, WT_3, HD_1, HD_2, HD_3)
```

Normalize and integrate data to remove batch effects (takes some time, feel free to skip and import pre-processed object)
```{r}
DefaultAssay(full_object) = 'RNA'

full_object <- full_object %>%
  NormalizeData() %>% FindVariableFeatures() %>% ScaleData() %>% RunPCA(reduction.name = "pca") 

full_object <- IntegrateLayers(
  object = full_object, method = CCAIntegration,
  orig.reduction = "pca", new.reduction = "cca")
```

Identify unbiased clusters
```{r}
full_object <- full_object %>% FindNeighbors(reduction = "cca", dims = 1:30) %>% FindClusters(cluster.name = "cca_clusters") 
```

Cells in clusters 11 and 17 are marked by ribosomal and mitochondrial gene expression, respectively, rather than cell type markers and are thus removed as bona-fide 'low quality cells'
```{r}
full_object = full_object %>% 
  subset(idents = c(11, 17), invert = TRUE) %>%
  FindNeighbors(reduction = "cca", dims = 1:30) %>% FindClusters(cluster.name = "cca_clusters") %>%
  RunUMAP(reduction = "cca", dims = 1:30, reduction.name = "umap.cca")

DimPlot(full_object, label = T, group.by = 'cca_clusters', reduction = 'umap.cca', split.by = 'condition') +coord_fixed() +NoAxes() + ggtitle('')
```

# Data analysis

Import pre-processed seurat object from GEO (RECOMMENDED)
```{r}
full_object= readRDS('../GEO_submission/full_object.rds')
```

Marker-based cluster annotation of main cell types (Supplementary Data 1)
```{r}
full_object= JoinLayers(full_object)
cluster_markers= FindAllMarkers(full_object, only.pos = T, min.pct = 0.25)

full_object <- RenameIdents(object = full_object, 
                         '0' = 'Progenitors', 
                         '1' = 'Mature neurons',
                         '2' = 'Mature neurons', 
                         '3' = 'Progenitors', 
                         '4' = 'Mature neurons', 
                         '5' = 'Maturing neurons', 
                         '6' = 'Mature neurons',
                         '7' = 'Mature neurons',  
                         '8' = 'Mature neurons', 
                         '9' = 'Mature neurons', 
                         '10'= 'Mature neurons',
                         '11'= 'Progenitors',
                         '12'= 'Mature neurons', 
                         '13'= 'Mature neurons', 
                         '14'= 'Mature neurons', 
                         '15'= 'Progenitors',
                         '16'='Proliferating progenitors',
                         '17'='Proliferating progenitors')

full_object$celltype_annotation <- Idents(full_object)
full_object$celltype_annotation= factor(as.character(full_object$celltype_annotation), levels=rev(c('Proliferating progenitors','Progenitors',  'Maturing neurons','Mature neurons')))
```

Generate plot for figure panels (Figures 2e-g and S3e-f)
```{r}
color_palette= c('Proliferating progenitors'= '#D9ED92','Progenitors'='#76C893',  'Maturing neurons'="#168AAD",'Mature neurons'="#184E77")

# Figure 2.e
DimPlot(full_object, label = F, group.by = 'celltype_annotation', reduction = 'umap.cca', split.by = 'condition', cols = color_palette) +coord_fixed() +NoAxes() + ggtitle('')

# Figure 2.f
ggplot(full_object@meta.data, aes(x= condition,fill=celltype_annotation)) + geom_bar(position='fill') + NoLegend()+RotatedAxis()+xlab('')+ scale_fill_manual(values = color_palette)

# Figure 2.g
FeaturePlot(full_object, c('PTPRZ1', 'TOP2A', 'ROBO3', 'MAPT'), reduction = 'umap.cca', max.cutoff = 1.5) *NoAxes()*coord_fixed()

# Figure S3.e
ggplot(full_object@meta.data, aes(x= sample_id,fill=celltype_annotation)) + geom_bar(position='fill') + NoLegend()+RotatedAxis()+xlab('')+ scale_fill_manual(values = color_palette)

# Figure S3.f
DimPlot(full_object, label = F, group.by = 'celltype_annotation', reduction = 'umap.cca', split.by = 'sample_id', ncol=3, cols = color_palette) + NoLegend() +coord_fixed() +NoAxes() + ggtitle('')
```